Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario

### 2021-04-28 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton-abstract.Rmd
Overall proofreading and editing of conclusions

---
title: "Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario"
author:
  - name: Author One
    email: author1@example.com
    affiliation: Some University
  - name: Author Two
    email: author2@example.com
    affiliation: Another University
    footnote: 1
  - name: Author Three
    email: author3@example.com
    affiliation: Some University
  - name: Author Four
    email: author4@example.com
    affiliation: Some University
address:
  - code: Some University
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  In this paper we analyze the changes in accessibility to food banks and related services before and during the COVID-19 pandemic in the City of Hamilton, Ontario. Food banks and services are the last line of support for households facing food insecurity; as such, they have become essential and their relevance cannot be ignored in the midst of the economic upheaval caused by the pandemic. Our analysis is based on the application of balanced floating catchment areas, and concentrates on households with lower incomes (<CAD40,000, approximately the Low Income Cutoff Value for a city of Hamilton's size). Accessibility was low to begin with in suburban and exurban parts of the city. We find that about 14\% of locations originally available in Hamilton closed during the pandemic; the impact of those closures on the level of service and accessibility was disproportionate, with system-wide losses exceeding 40\%. Those losses were geographically and demographically uneven. While every part of the city faced a reduction in accessibility, inner suburbs fared worse in terms of loss of accessibility. As well, children (age $\le 18$) appear to have been impacted the most.

>>>>

---
title: "Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario"
author:
  - name: Author One
    email: author1@example.com
    affiliation: Some University
  - name: Author Two
    email: author2@example.com
    affiliation: Another University
    footnote: 1
  - name: Author Three
    email: author3@example.com
    affiliation: Some University
  - name: Author Four
    email: author4@example.com
    affiliation: Some University
address:
  - code: Some University
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  In this paper we analyze the changes in accessibility to food banks and related services before and during the COVID-19 pandemic in the City of Hamilton, Ontario. Food banks and services are the last line of support for households facing food insecurity; as such, their relevance cannot be ignored in the midst of the economic upheaval caused by the pandemic. Our analysis is based on the application of balanced floating catchment areas, and concentrates on households with lower incomes (<CAD40,000, approximately the Low Income Cutoff Value for a city of Hamilton's size). We find that accessibility was low to begin with in suburban and exurban parts of the city; furthermore, about 14\% of locations originally available in Hamilton closed during the pandemic, further reducing accessibility. The impact of closures on the level of service of the remaining facilities, and on accessibility, was disproportionate, with system-wide losses exceeding 40\%. Those losses were geographically and demographically uneven. While every part of the city faced a reduction in accessibility, inner suburbs fared worse in terms of loss of accessibility. As well, children (age $\le 18$) appear to have been impacted the most.

### 2021-04-28 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton-title-page.Rmd
Overall proofreading and editing of conclusions

---
title: "Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario"
author:
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: University of Toronto Scarborough
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Gyoorie Kim
    email: gyoorie.kim@mail.utoronto.ca
    affiliation: University of Toronto Mississauga
  - name: Jue Wang
    email: gis.wang@utoronto.ca
    affiliation: University of Toronto Mississauga
address:
  - code: University of Toronto Scarborough
    address: Department of Geography & Planning, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON M1C1A4
  - code: McMaster University
    address: School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada
  - code: University of Toronto Mississauga
    address: Department of Geography, Geomatics and Environment, University of Toronto Mississauga, 3359 Mississauga Road, Mississauga, ON, L5L 1C6 , Canada   
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  In this paper we analyze the changes in accessibility to food banks and related services before and during the COVID-19 pandemic in the City of Hamilton, Ontario. Food banks and services are the last line of support for households facing food insecurity; as such, they have become essential and their relevance cannot be ignored in the midst of the economic upheaval caused by the pandemic. Our analysis is based on the application of balanced floating catchment areas, and concentrates on households with lower incomes ($<CAD40,000, approximately the Low Income Cutoff Value for a city of Hamilton's size). Accessibility was low to begin with in suburban and exurban parts of the city. We find that about 14\% of locations originally available in Hamilton closed during the pandemic; the impact of those closures on the level of service and accessibility was disproportionate, with system-wide losses exceeding 40\%. Those losses were geographically and demographically uneven. While every part of the city faced a reduction in accessibility, inner suburbs fared worse in terms of loss of accessibility. As well, children (age $\le 18$) appear to have been impacted the most.

>>>>

---
title: "Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario"
author:
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: University of Toronto Scarborough
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Gyoorie Kim
    email: gyoorie.kim@mail.utoronto.ca
    affiliation: University of Toronto Mississauga
  - name: Jue Wang
    email: gis.wang@utoronto.ca
    affiliation: University of Toronto Mississauga
address:
  - code: University of Toronto Scarborough
    address: Department of Geography & Planning, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON M1C1A4
  - code: McMaster University
    address: School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada
  - code: University of Toronto Mississauga
    address: Department of Geography, Geomatics and Environment, University of Toronto Mississauga, 3359 Mississauga Road, Mississauga, ON, L5L 1C6 , Canada   
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  In this paper we analyze the changes in accessibility to food banks and related services before and during the COVID-19 pandemic in the City of Hamilton, Ontario. Food banks and services are the last line of support for households facing food insecurity; as such, their relevance cannot be ignored in the midst of the economic upheaval caused by the pandemic. Our analysis is based on the application of balanced floating catchment areas, and concentrates on households with lower incomes (<CAD40,000, approximately the Low Income Cutoff Value for a city of Hamilton's size). We find that accessibility was low to begin with in suburban and exurban parts of the city; furthermore, about 14\% of locations originally available in Hamilton closed during the pandemic, further reducing accessibility. The impact of closures on the level of service of the remaining facilities, and on accessibility, was disproportionate, with system-wide losses exceeding 40\%. Those losses were geographically and demographically uneven. While every part of the city faced a reduction in accessibility, inner suburbs fared worse in terms of loss of accessibility. As well, children (age $\le 18$) appear to have been impacted the most.

### 2021-04-28 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd
Overall proofreading and editing of conclusions

Food insecurity is defined as an "inadequate or uncertain access to a sufficient quantity and/or adequate quality of food" due to a household's financial limitations [@enns2020experiences]. This condition has been associated with reductions in nutritional outcomes [@bhattacharya2004poverty; @kirkpatrick2008food; @olson1999nutrition] and physical and mental health in children and adults [@elgar2021relative; @jones2017food; @ramsey2011food; @seligman2010food; @stuff2004household]. Over at least the past four decades food banks and related services have become an essential line of defense against food insecurity in Canadian communities [@riches2002food; @holmes2018nothing; @black2020examining; @tarasuk2020relationship]. In this respect, Canada is not unlike numerous other wealthy countries where a systematic dismantling of the welfare state took place in the intervening period [@tarasuk2014food].

The emergence of COVID-19, the worst public health crisis since the 1918 flu pandemic, has revealed important social and economic fault lines, and pre-existing patterns of inequality appear to have been exacerbated. Along several other dimensions of stress [e.g., accessibility to health care facilities, @pereira2021geographic], this seems to be the case for food insecurity as well [@laborde2020poverty]. In the US, for example, it has been estimated that there was an increase of more than 30\% in household food insecurity, and more than one third of households were discovered to be newly food insecure - meaning they did not experience food insecurity before the pandemic [@niles2020early]. In Canada, Men and Tarasuk [-@men2021food] report that about 25\% of individuals who experienced job insecurity (a relatively common occurrence during the pandemic), also experienced food insecurity. Similarly, according to Statistics Canada [-@statisticscanada2020food], in the early stages of the pandemic almost 15\% of individuals reported living in a household that faced food insecurity; the risk of food insecurity was substantially higher for households with children. The difference between households with and without children was significant, and 11.7\% of households with children indicated that "food didn’t last and [there was] no money to get more" sometimes or often, compared to 7.3\% of households without children; likewise, 13\% of households with children indicated that they "[c]ouldn’t afford balanced meals" sometimes or often, compared to 8.8\% of households without children.

@@

Economic stress caused by the COVID-19 pandemic, rising unemployment rates, and changes in poverty levels have disrupted the food environment and led to higher rates of food insecurity [@niles2020early]. Pandemic-related food security studies in the US have found a substantial increase in households experiencing food insecurity for the first time, and also in households experiencing more severe food insecurity than before [@niles2020early; @wolfson2020food]. Most recently in May 2020, Canada recorded 14.7% of its population living in food insecurity in the past 30 days [@statisticscanada2020food]. Considering the negative mental and physical health effects associated with food insecurity, increases in food insecurity rates due to the pandemic, signal a change in the food environment with potential damages to population health during the course of the pandemic and beyond [@niles2020early]. Recent data^[https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis_Exec-Sum_EN.pdf] from FBC showed that 52% of member food banks reported an increase in usage in March of 2020 when initial lockdown restrictions were put in place across much of the country. The pandemic also created significant staffing issues with 42% of food banks reporting a reduction in volunteers. However, 53% of food banks later reported a decrease in use into the summer of 2020 which FBC members attributed to emergency financial support programs from the federal government. Nevertheless, some of these benefit programs were temporary. Although more recent statistics on food bank use in Hamilton in 2020 and 2021 is not yet available, data from the Daily Bread Food Bank in neighbouring Toronto for August 2020 shows visits climbing 51% year-over-year in that city^[https://www.dailybread.ca/wp-content/uploads/2020/11/Whos-Hungry-Report-2020.pdf], which suggests that many households in Hamilton are likely to turn to food bank services to meet their needs.

@@

With a population of around 540,000, the City of Hamilton is the fourth largest city in Ontario. It has historically been home to major manufacturing industries but de-industrialization that has occurred over the past several decades has led Hamilton to become one of the most highly divided cities in Ontario, with a significant proportion of its residents sustaining livings at Canada’s poverty level [@deluca2012code; @jakar2019turning; @latham2007determinants]. In 2016, the Hamilton Community Foundation reported that based on the Low-Income Cut-Off, Hamilton recorded a poverty rate of 16.7%, which was well above the average rate of Ontario (13.7%) and the average national rate (12.8%).^[https://www.hamiltoncommunityfoundation.ca/vital-signs/low-income-2018/]. According to data from Hamilton Food Share^[https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf], approximately 23,000 individuals accessed food banks in the city in March of 2019. Within this total is 9,125 visits by children (minors up to 18 years old), up from 8,278 the year before. Feed Ontario, the province's largest collective of hunger-relief organizations, found that on a per-capita basis, the level of need in the inner core of central Hamilton was second highest in Ontario^[https://feedontario.ca/research/report-hunger-map/].

@@

Figure \ref{fig:sensitivity-analysis} shows the results of conducting sensitivity analysis of the system-wide accessibility as we vary the threshold (considering the situation before the pandemic). There is a clear pattern whereby more strict values of $\delta$ are associated with higher levels of system-wide accessibility: while increases in accessibility that result from decreases in the travel time window might seem counter-intuitive, this is a result of lower _congestion_, since fewer households are serviced and thus competition for the same resources is more limited. System-wide accessibility declines with higher values of $\delta$: as more households are serviced, congestion grows and the level of service declines, although this happens at a declining rate. We are not aware of any research that explains how long people are expected to travel for food banks, but we note that in developing countries, access to drinking water is defined as a source that takes less than 30 minutes to reach [round trip, see @world2019progress]. There is no reason why people in affluent countries should be expected to travel more for a basic need such as food. Accordingly, we adopt a 15-minute threshold for the analysis (representing a one-way trip). This threshold is also approximately where the rate of change in accessibility slows down.

@@

We find that the accessibility landscape deteriorated substantially during the pandemic, with accessibility dropping on average by almost `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% mean() %>% abs() %>% round()`\%, but with large variations: some zones experienced changes in accessibility of only about `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% max() %>% abs() %>% round()`\%, whereas the most affected zone saw a loss of accessibility of almost `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% min() %>% abs() %>% round()`\%. Figure \ref{fig:accessibility-changes-with-local-i} shows the changes in accessibility. Every zone is worse off after the closure of facilities during the pandemic, but some parts of the city seem to have been particularly affected. To better highlight these changes, we used a local indicator of spatial autocorrelation [@anselin1995local] to explore the pattern of change in accessibility. Twenty-four TAZ are flagged as having significant large losses of accessibility (at $p<=0.10$, without correcting for multiple comparisons). Those zones are highlighted in the figure, where it can be seen that they form more or less compact neighborhoods. Remarkably, the largest (and significant) drops in accessibility are not downtown, but located in two cases in the industrial north of the city, in one case in an inner suburb above the escarpment, and lastly in a more suburban/exurban region in the south-west.

@@

Just as the effects of the closures appear to have been uneven in space, they also seem to have differently impacted various population segments. Table \ref{tab:accessibility-by-age} shows the estimated number of people in each age group by their level of accessibility before and during the pandemic. Here, it is important to note that the quartiles are relative: people in the top 25\% of accessibility still have lower accessibility during the pandemic than before. In reality, every population group is worse off during the pandemic in terms of their accessibility to food banks and services in the City of Hamilton. However, some age groups were affected more. As seen in the table, the distribution within quartiles for adults changed only slightly, and in fact there are fewer adults in the bottom three quartiles as some found themselves in the (deteriorated) top quartile of accessibility during the pandemic. The story is different for seniors, a greater number of whom are now in the third quartile, due to a combination of seniors facing a worse accessibility situation, and just a few moving from the bottom quartile (pre-pandemic) to the third during the pandemic. Those aged 18 and less appear to have fared worse, and the largest change is in the number of children who were in the third quartile before the pandemic and found themselves in the bottom accessibility class during the pandemic. These results suggest that, through a combination of the typical modes of transportation of lower income households and the spatial distribution of the population, the food bank/service locations closed had a differential impact that more greatly affected the youngest among the population in low income households.

@@

Food insecurity is a significant issue for many Canadian households and is associated reductions in nutrition and mental and physical health. While food bank use can provide some relief, levels of need in the City of Hamilton were on the rise before the onset of the COVID-19 pandemic with a troubling increase in the number of visits made by children. While temporary financial relief measures helped reduce food bank use in the summer of 2020, the pandemic has more than likely increased food insecurity for many households and created a surge in food bank demand in the city. On the other hand, the pandemic has also resulted in major disruptions to food bank and food service delivery, resulting in the closure or cancellation of many food services in Hamilton. On top of this, reductions in transit service have reduced travel options for many.

In response, this research has sought to better understand accessibility to food banks and food service locations as well as how the closure of some locations over the pre- and during-COVID time periods affected the potential for low-income households to reach these amenities. Using the balanced floating catchment area method to account for population demand and congestion effects at service points, we estimate multi-modal accessibility to food bank and food service locations for low-income households. In the absence of information on how food insecure households travel to food banks and related services, we adopted a normative approach that examines accessibility to food banks reachable within 15 minutes based on global standards for accessible drinking water. The weighting of travel time estimates by the modal split in different zonal geographies tailors the results to patterns of travel behaviour captured in the TTS regional travel survey. Moreover, our parcel-level analysis presents a disaggregate approach to estimating accessibility based on the locations of residential parcels and dwellings.

Results show that while accessibility levels were lower in the city's more car-oriented suburban and rural areas to begin with, the closure of 14% of the city's food bank and low-cost or free food service locations during the pandemic resulted in an overall decrease in accessibility across the city. However, these effects were not uniform over space or for different population groups. Because the balanced floating catchment area method takes into account changes in demand and congestion for service providers, the closure of food bank and food service locations can reverberate throughout the catchment areas of the city. For some suburban zones, the closure of a relatively high level-of-service location results in the remaining services being spread over a larger population. In others, high auto dependence for trips leads to decreases in accessibility that accumulate due to the loss of several food bank and service locations initially reachable within 15 minutes by car. Reductions in accessibility in the city's more urban north end, where low-income households conduct higher proportions of trips by transit and walking, emphasize the importance of geographic proximity in the potential to reach food bank and service locations for these residents. Beyond geography, the results also highlight the differential impact of food bank closing during COVID-19 on population groups with seniors and children most affected.

Still, the work is not without its limitations. We use data from the census to model low-income households as a proxy for households that are possibly experiencing food insecurity. However, the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known. We also consider all food banks and food service locations equally. Future work could improve this method by accounting for differential capacities at food provider locations, such as the number of meals served. Unforunately, this data was not available in the present case as some service locations had already closed. We also did not explicitly account for any changes in demand over the study period. If trends in Hamilton are similar to those in Toronto, there could be a significant increase in demand that would alter the estimated accessibilities. Moreover, while the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown as approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations.

Nevertheless, while previous work has noted the important role of geography alongside other socioeconomic and demographic indicators in household access to healthy food, this is, to the best of our knowledge, the first study that has focused on the geographic component of food bank accessibility. Beyond accounting for the inflation of demand that occurs in traditional floating catchment area methods, our application of the balanced floating catchment area approach offers a novel analysis of food bank and food service accessibility that is sensitive to the locations of low-income households, information on their age distributions and typical tripmaking behaviour, food bank and food service locations and their operations over time, and the characteristics of the city's multi-modal transportation network. Findings from this analysis can provide quantitative evidence to help urban planners and community administrators strategically allocate food bank services considering accessibility for different population groups, especially when considering the unequal distribution of impacts from the COVID-19 pandemic.

>>>>

Food insecurity is defined as an "inadequate or uncertain access to a sufficient quantity and/or adequate quality of food" due to a household's financial limitations [@enns2020experiences]. This condition has been associated with reductions in nutritional outcomes [@bhattacharya2004poverty; @kirkpatrick2008food; @olson1999nutrition] and negative physical and mental health impact in children and adults [@elgar2021relative; @jones2017food; @ramsey2011food; @seligman2010food; @stuff2004household]. Over at least the past four decades food banks and related services have become an essential line of defense against food insecurity in Canadian communities [@riches2002food; @holmes2018nothing; @black2020examining; @tarasuk2020relationship]. In this respect, Canada is not unlike numerous other wealthy countries where a systematic dismantling of the welfare state took place in the intervening period [@tarasuk2014food].

The emergence of COVID-19, the worst public health crisis since the 1918 flu pandemic, has exposed important social and economic fault lines, and pre-existing patterns of inequality appear to have been exacerbated. Along several other dimensions of stress [e.g., accessibility to health care facilities, @ghorbanzadeh2021spatial; @pereira2021geographic], this seems to be the case for food insecurity as well [@laborde2020poverty]. In the US, for example, it has been estimated that there was an increase of more than 30\% in household food insecurity, and more than one third of households were discovered to be newly food insecure - meaning they did not experience food insecurity before the pandemic [@niles2020early]. In Canada, Men and Tarasuk [-@men2021food] report that about 25\% of individuals who experienced job insecurity (a relatively common occurrence during the pandemic), also experienced food insecurity. Similarly, according to Statistics Canada [-@statisticscanada2020food], in the early stages of the pandemic almost 15\% of individuals reported living in a household that faced food insecurity; the risk of food insecurity was substantially higher for households with children. The difference between households with and without children was significant, and 11.7\% of households with children indicated that "food didn’t last and [there was] no money to get more" sometimes or often, compared to 7.3\% of households without children; likewise, 13\% of households with children indicated that they "[c]ouldn’t afford balanced meals" sometimes or often, compared to 8.8\% of households without children.

@@

Economic stress caused by the COVID-19 pandemic, rising unemployment rates, and changes in poverty levels have disrupted the food environment and led to higher rates of food insecurity [@niles2020early]. Pandemic-related food security studies in the US have found a substantial increase in households experiencing food insecurity for the first time, and also in households experiencing more severe food insecurity than before [@niles2020early; @wolfson2020food]. Most recently in May 2020, Canada recorded 14.7% of its population living in food insecurity in the past 30 days [@statisticscanada2020food]. Considering the negative mental and physical health effects associated with food insecurity, increases in food insecurity rates due to the pandemic, signal a change in the food environment with potential damages to population health during the course of the pandemic and beyond [@niles2020early]. Recent data^[https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis_Exec-Sum_EN.pdf] from FBC showed that 52% of member food banks reported an increase in usage in March of 2020 when initial lockdown restrictions were put in place across much of the country. The pandemic also created significant staffing issues with 42% of food banks reporting a reduction in volunteers. However, 53% of food banks later reported a decrease in use into the summer of 2020 which FBC members attributed to emergency financial support programs from the federal government. Nevertheless, some of these benefit programs were temporary. Although more recent statistics on food bank use in Hamilton in 2020 and 2021 are not yet available, data from the Daily Bread Food Bank in neighbouring Toronto for August 2020 shows visits climbing 51% year-over-year in that city^[https://www.dailybread.ca/wp-content/uploads/2020/11/Whos-Hungry-Report-2020.pdf], which suggests that many households in Hamilton are likely to turn to food bank services to meet their needs.

@@

With a population of around 540,000, the City of Hamilton is the fourth largest city in Ontario. It has historically been home to major manufacturing industries but de-industrialization that has occurred over the past several decades has led Hamilton to become one of the most highly divided cities in Ontario, with a significant proportion of its residents living at or below Canada’s poverty level [@deluca2012code; @jakar2019turning; @latham2007determinants]. In 2016, the Hamilton Community Foundation reported that based on the Low-Income Cut-Off, Hamilton recorded a poverty rate of 16.7%, which was well above the average rate of Ontario (13.7%) and the average national rate (12.8%).^[https://www.hamiltoncommunityfoundation.ca/vital-signs/low-income-2018/]. According to data from Hamilton Food Share^[https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf], approximately 23,000 individuals accessed food banks in the city in March of 2019. Within this total is 9,125 visits by children (minors up to 18 years old), up from 8,278 the year before. Feed Ontario, the province's largest collective of hunger-relief organizations, found that on a per-capita basis, the level of need in the inner core of central Hamilton was second highest in Ontario^[https://feedontario.ca/research/report-hunger-map/].

@@

Figure \ref{fig:sensitivity-analysis} shows the results of conducting sensitivity analysis of the system-wide accessibility as we vary the threshold (considering the situation before the pandemic). There is a clear pattern whereby more strict values of $\delta$ are associated with higher levels of system-wide accessibility: while increases in accessibility that result from decreases in the travel time window might seem counter-intuitive, this is a result of lower _congestion_, since fewer households are serviced and thus competition for the same resources is more limited. System-wide accessibility declines with higher values of $\delta$: as more households are serviced, congestion grows and the level of service declines, although this happens at a declining rate. We are not aware of any research that explains how long people are expected to travel for food banks, but we note that in developing countries, accessible sources of drinking water are those that can be reached in less than 30 minutes [round trip, see @world2019progress]. There is no reason why people in affluent countries should be expected to spend more time travelling for a basic necessity such as food. Accordingly, we adopt a 15-minute threshold for the analysis (representing a one-way trip). This threshold is also approximately where the rate of change in accessibility slows down.

@@

We find that the accessibility landscape deteriorated substantially during the pandemic, with accessibility dropping on average by almost `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% mean() %>% abs() %>% round()`\%, but with large variations: some zones experienced changes in accessibility of only about `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% max() %>% abs() %>% round()`\%, whereas the most affected zone saw a loss of accessibility of almost `r accessibility_taz_change %>% st_drop_geometry() %>% summarize(d = (.[[6]] - .[[5]])/.[[5]] * 100) %>% pull(d) %>% min() %>% abs() %>% round()`\%. Figure \ref{fig:accessibility-changes-with-local-i} shows the changes in accessibility. Every zone is worse off after the closure of facilities during the pandemic, but some parts of the city seem to have been particularly affected. To better highlight these changes, we used a local indicator of spatial autocorrelation [@anselin1995local] to explore the pattern of change in accessibility. Twenty-four TAZ are flagged as having significantly large losses of accessibility (at $p\le 0.10$, without correcting for multiple comparisons). Those zones are highlighted in the figure, where it can be seen that they form more or less compact neighborhoods. Remarkably, the largest significant drops in accessibility are not downtown, but located in two cases in the industrial north of the city, in one case in an inner suburb above the escarpment, and lastly in a more suburban/exurban region in the south-west.

@@

Just as the effects of the closures appear to have been uneven in space, they also seem to have had different impacts on various population segments. Table \ref{tab:accessibility-by-age} shows the estimated number of people in each age group by their level of accessibility before and during the pandemic. Here, it is important to note that the quartiles are relative: people in the top 25\% of accessibility still have lower accessibility during the pandemic than before. In reality, every population group is worse off during the pandemic in terms of their accessibility to food banks and services in the City of Hamilton. However, some age groups were affected more. As seen in the table, the distribution within quartiles for adults changed only slightly, and in fact there are fewer adults in the bottom three quartiles as some found themselves in the (deteriorated) top quartile of accessibility during the pandemic. The story is different for seniors, a greater number of whom are now in the third quartile, due to a combination of seniors facing a worse accessibility situation, and just a few moving from the bottom quartile (pre-pandemic) to the third during the pandemic. Those aged 18 and less appear to have fared worse, and the largest change is in the number of children who were in the third quartile before the pandemic and found themselves in the bottom accessibility class during the pandemic. These results suggest that, through a combination of the typical modes of transportation of lower income households and the spatial distribution of the population, the food bank/service locations closed had a differential impact that more greatly affected the youngest among the population in low income households.

@@

Food insecurity is a significant issue for many Canadian households and is associated with reductions in nutrition and poor mental and physical health outcomes. While food bank use can provide some relief, the levels of need in the City of Hamilton were on the rise before the onset of the COVID-19 pandemic, with a troubling increase in the number of visits made by children. While temporary financial relief measures may have helped reduce food bank use in the summer of 2020, the pandemic has in all probability increased food insecurity for many households. To compound matters, the pandemic has also resulted in major disruptions, including to employment, mobility alternatives, and to food bank and food service delivery. In response, this research has sought to better understand accessibility to food banks and food service locations, as well as how the closure of some locations over the pre- and during-COVID time periods affected the potential for low-income households to reach these amenities. Using the balanced floating catchment area method to account for population demand and congestion effects at service points, we estimated multi-modal accessibility to food bank and food service locations for low-income households. The weighting of travel time estimates by the modal split in different zonal geographies tailors the results to patterns of travel behaviour captured in the regional travel survey. Moreover, our parcel-level analysis presents a disaggregate approach to estimating accessibility based on the locations of residential parcels and dwellings.

Our results show that while accessibility levels were lower in the city's more car-oriented suburban and rural areas to begin with, the closure of 14% of the city's food bank and low-cost or free food service locations during the pandemic resulted in an overall decrease in accessibility across the city. However, these effects were not uniform over space or for different population groups. Since the balanced floating catchment area method takes into account changes in demand and congestion for service providers, the closure of food bank and food service locations can reverberate throughout the catchment areas of the whole city. For some suburban zones, the closure of a relatively high level-of-service location results in the remaining services being spread over a larger population. In others, high auto dependence for trips leads to decreases in accessibility that accumulate due to the loss of several food bank and service locations initially reachable within 15 minutes by car. Reductions in accessibility in the city's more urban north end, where low-income households conduct higher proportions of trips by transit and walking, emphasize the importance of geographic proximity in the potential to reach food bank and service locations for these residents. Beyond geography, the results also highlight the differential impact of food banks closing during COVID-19 on population groups, with seniors and children most affected.

In the absence of information regarding how food insecure households travel to food banks and related services, we examined accessibility to food banks using a 15 minutes (one way) travel time threshold. The fact that we must rely on a standard created for accessible drinking water in the developing world only serves to highlight the tragedy of food insecurity in an affluent country like Canada. More broadly, it points to the absurd need to understand how a bad situation was made worse by the pandemic: in effect, the analysis reveals that disparities in accessibility to food banks and services predated the pandemic, that the pandemic contributed to the deterioration of these services, and that populations already in distress, particularly children, ended up in an even more adverse state. How much worse, it is impossible to say, mainly because there is also a dearth of information, let alone standards, regarding what is an _acceptable_ or _sufficient_ level of service when it comes to food banks. Whether this is one facility per 1,000 population or three facilities per 1,000 population, or any other number, we know not.

In summary, while previous work has noted the important role of geography alongside other socio-economic and demographic indicators in household access to healthy food, this is, to the best of our knowledge, the first study that has focused on the geographic component of food bank accessibility. Beyond accounting for the inflation of demand that occurs in traditional floating catchment area methods, our application of the balanced floating catchment area approach offers a novel analysis of food bank and food service accessibility that is sensitive to the locations of low-income households, information on their age distributions and typical trip-making behaviour, food bank and food service locations and their operations over time, and the characteristics of the city's multi-modal transportation network. Despite the limitations noted above, the findings from this analysis can provide quantitative evidence to help urban planners and community administrators strategically allocate food bank services considering accessibility for different population groups, especially when considering the unequal distribution of impacts from the COVID-19 pandemic.

In terms of future research, it is important to note that the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known. We also consider all food banks and food service locations equally. Information on differential capacities at food provider locations is currently not collected, and given the closure of facilities was not possible to obtain. Data on services, such as the number of meals served, could be used to refine the analysis in the future. Moreover, while the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown, and approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations, and our estimates might be somewhat conservative if those households who do not report income rely more on walking and/or transit for their mobility needs.
